Designing Types for R, Empirically

### 2020-05-16 paper/main.tex
intro

\begin{abstract}
The R programming language is widely used in a variety of domains for tasks
related to data science. R was designed to favor an interactive style of
programming with minimal syntactic and conceptual overhead. This design is
well suited to interactive data analysis, but a bad fit for tools such as
compilers or program analyzers which must generate performant code or catch
programming errors.  In particular, R has no type annotations, its
operations are dynamically checked at run-time. The starting point for our
work are the twin questions, \emph{what expressive power is needed to
  accurately type R code?} and \emph{which type system is the R community
  willing to adopt?} Both questions are difficult to answer without actually
designing a type system and attempting to convince users and package
developers to try the proposed system.  The goal of this paper is to provide
data that can feed into the design process. To this end, we perform a large
corpus analysis with aim to gain insights in the degree of polymorphism
exhibited by idiomatic R code and explore potential benefits that the R
community could accrue from even a simple type system.  As a starting point,
we infer type signatures for \NUMFUNCTIONS functions from \NUMPACKAGES
packages among the most widely used open source R libraries. The signatures
use a minimal type language with only basic types, vectors and classes.  We
implement a contract checker that can be used to validate the inferred type
signatures.
\end{abstract} \maketitle

@@

In this section, we will introduce related work and the R programming language.

@@

There is no previous work for the R programming language. We take
inspiration in the above mentioned works but focus on adapting them to our
target language.

>>>>

\begin{abstract}
The R programming language is widely used in a variety of domains.  It was
designed to favor an interactive style of programming with minimal syntactic
and conceptual overhead. This design is well suited to interactive data
analysis, but a bad fit for tools such as compilers or program analyzers
which must generate native code or catch programming errors.  In particular,
R has no type annotations, all operations are dynamically checked at
run-time. The starting point for our work are the twin questions, \emph{what
  expressive power is needed to accurately type R code?} and \emph{which
  type system is the R community willing to adopt?} Both questions are
difficult to answer without actually experimenting with a type system. The
goal of this paper is to provide data that can feed into that design
process. To this end, we perform a large corpus analysis to gain insights in
the degree of polymorphism exhibited by idiomatic R code and explore
potential benefits that the R community could accrue from a simple type
system.  As a starting point, we infer type signatures for \NUMFUNCTIONS
functions from \NUMPACKAGES packages among the most widely used open source
R libraries.
\end{abstract} \maketitle

@@

In this section, we introduce related work on extending dynamic languages
with static type system and we give a short primer on the R programming
language.

@@

There is no previous work on types for the R programming language. We take
inspiration in the above mentioned works but focus on adapting them to our
target language.
